Split-Apply-Combine with Dynamic Grouping

### 2023-01-05 paper/article.Rnw
writing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extensive example: economic data} \label{sect:example}

\section{Formal descriptions and algorithms}
In this section we give a formal description of the algorithms. In order to be
unambiguous and to represent the algorithms compactly, we introduce some
notation. The notation also makes the pseudocode independent of the
implementation language used, or how the data is represented.

We start by giving the algorithm for ordinary split-apply-combine, so it is
easy to see where the algorithm must be generalized to allow for a collapsing
scheme.

@@

Let $U$ be a finite set, and let $\phi$ be a function that accepts a subset of
$U$ and returns a value in some domain $X$. Here, $U$ represents a data set and
$\phi$ an aggregating function. We split $U$ into groups using the following
notation. Let $A$ be finite set that has no more elements than $U$, and let $f$
be a function that takes an element of $U$ and returns a value in $A$. We can
think of $A$ as a set of group labels, and $f$ as the function that assigns a
label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
groups, since every element in $U$ can only get one label.  In formal
terminology we say that $f:U\to A$ is a \emph{partition} of $U$.

@@

\begin{algorithm}[H]
\caption{Split-Apply-Combine}
\label{alg:sac}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition $f:U\to A$}
\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs $(a,x)\in
A\times X$ }

$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}

Here, the output is collected in a set $R$ (this set is technically a function
$A\to X$ since it pairs for each $a\in A$ with one element of $X$). We also
introduced some new notation: the notation $2^U$ stands for `the set of all
subsets of $U$'. The notation $f^{-1}$ stands for the formal \emph{inverse} of
$f$. That is: you give it a set of labels from $A$ and it returns all the
elements of $U$ that have any of those labels. This definition is slightly too
general for the current algorithm, but it will come in handy when we move to
collapsing groups.

@@

Collapsing schemes can be represented as tables. For example, the European NACE
classification is a hierarchical classification of economic activity. Each
group is encoded as a 4-digit code where the first digit indicates the highest
grouping level, and the last digit the most detailed level. This induces a
collapsing sequence by combining groups that have the same parent in the
hierarchy. Below is an example, based on a small piece of the NACE code list.
%
\begin{equation}
\begin{tabular}{rrr}
\hline
 $A_0$& $A_1$ & $A_2$\\
\hline
0111&   011 & 01\\
0112&   011 & 01\\
0113&   011 & 01\\
0121&   012 & 01\\
0122&   012 & 01\\
0123&   012 & 01\\
0124&   012 & 01\\
\hline
\end{tabular}
\label{eq:tablerep}
\end{equation}

Here, $A_0$ is the most detailed partition, splitting a dataset into seven
groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
$0113$ and the second, labeled $01.2$ is constructed by combining
$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.

To compute an aggregate for records labeled with $0111$, the algorithm will
select those records and test whether the subset is suitable. If not, it will
move to $f_1(0111)=011$ and find all the groups in $A_0$ that map to 011 as
follows.
\begin{displaymath}
f_1^{-1}(\{011\}) = \{0111, 0112, 0113\}.
\end{displaymath}
It will then move on and find all records labeled with any of those three
group labels, and test again. If this new data set is suitable, the result is
computed, if not, another collapse takes place. The whole procedure is repeated 
for each element in $A_0$.

In order to generalize this algorithm towards generic collapsing schemes, we
introduce the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and
returns the corresponding label in $A_k$. This can formally be defined as
consecutive application of $f_1, f_2,\cdots, f_k$ to an element of $A_0$. In
notation
\begin{equation*}
F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
\end{equation*}
Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
\begin{equation*}
F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
\end{equation*}
This function asks for a set of labels in $A_k$ and returns all the labels in
$A_0$ that are mapped to those labels via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally write
down our main Algorithm~\ref{alg:saccg}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Extensive example: economic data} \label{sect:example}
In this section we treat three practical examples using a synthetic
dataset included with the package.
<<>>=
data(producers)
head(producers)
@
This \code{producers} dataset contains synthetic data records of various income
sources for \Sexpr{nrow(producers)} industrial produces. The records are
classified into a local version of the NACE classification called \code{sbi}
and into \code{size} classes with values in $\{5,6,7,8,9\}$.

\subsection{Small area estimation} 
Small area estimation (SAE) is a collection of methods that are aimed at
estimating subpopulation parameters in cases where there the number of
observations in a subpopulation is so mall that direct estimation leads to
unacceptable estimation variance. Instead one resorts to indirect estimation,
meaning that one estimates the parameter for a larger subpopulation that
includes the target subpopulation. This estimate is then interpreted as an
estimate for the target subpopulation. Here, we shall be interested in
estimating the average turnover from industrial activities (\code{industrial})
by SBI and size class.

In the simplest case, where no auxiliary information is available or used, one
replaces the estimator of the mean over a subpopulation with the estimator of
the mean over a larger subpopulation. If we assume that all units in the
dataset were sampled with equal probability based on a simple random sampling
design, the mean is estimated with the sample mean. We will demand that there
are at least ten records for which turnover has been measured. The collapsing
scheme is given by \code{sbi*size ~ sbi + sbi2 + sbi1} where \code{sbi2} and
\code{sbi1} are classification by respectively the first two SBI digits and the
first SBI digit. We first add those variables to the dataset.
<<>>=
producers$sbi2 <- substr(producers$sbi,1,2)
producers$sbi1 <- substr(producers$sbi,1,1)
head(producers, 3)
@
Our test function looks as follows.
<<>>=
testfun <- function(d) sum(!is.na(d$industrial)) >= 10
@
Using \code{accumulate()} we obtain the means. 
<<>>=
a <- cumulate(producers
             , collapse = sbi*size ~ sbi + sbi2 + sbi1
             , test = testfun
             , mean_industrial = mean(industrial, na.rm=TRUE))
head(a,3)
@

\subsection{Imputing missing values using SAE and ratio imputation} 
Our goal in this example is to impute missing values for the \code{industrial}
variable based on ratio imputation with \code{total} as predictor.  Ratio
imputation is a method where the imputed value $\hat{y}_i$ for variable $Y$ of
record $i$ is estimated as $\hat{y_i}=\hat{R}_d x_i$, where $\hat{R}$ is an
estimate of the ratio between $Y$ and an auxiliary variable $X$ in
subpopulation $d$. An unbiased estimate for $R_d$ is given by
$\hat{\bar{Y}}_d/\hat{\bar{X}}_d$ where $\hat{\bar{Y}}_d$ and $\hat{\bar{X}}_d$
are estimated subpopulation means.

We use SAE to estimate the subpopulation ratios, and then use the \code{simputation}
package \citep{loo2022simputation} to impute the missing values. 
<<>>=
r <- cumulate(producers
         , collapse = sbi*size ~ sbi + sbi2 + sbi1
         , test = testfun
         , R = mean(industrial, na.rm=TRUE)/mean(total, na.rm=TRUE))
head(r,3)
@
To impute the values, using \code{impute_proxy()} we need to merge the ratios
with the producers dataset (which automatically happens by SBI and size class).
<<>>=
library(simputation)
dat <- merge(producers, r)
dat <- impute_proxy(dat, industrial ~ R*total)
@
We can inspect the imputed values as follows.
<<>>=
iNA <- is.na(producers$industrial)
head(dat[iNA, c("sbi","size","level","R","industrial","total")])
@
From this, we have all the information to interpret the imputed values. For
example, we see that in record 664, the ratio was estimated after collapsing
the group $(\texttt{sbi},\texttt{size})=(21122,7)$ to $\texttt{sbi2}=21$ since
the collapse \code{level} equals 2.

\section{Formal descriptions and algorithms}
In this section we give a formal description of the algorithm for aggregation
with dynamic grouping.  We start by giving an algorithm for ordinary
split-apply-combine to show where the algorithm must be generalized to allow
for a collapsing scheme.

@@

Let $U$ be a finite set, and let $\phi$ be a function that accepts a subset of
$U$ and returns a value in some domain $X$. Here, $U$ represents a data set and
$\phi$ an aggregating function. We split $U$ into groups using the following
notation. Let $A$ be finite set that has no more elements than $U$, and let $f$
be a function that takes an element of $U$ and returns a value in $A$. We can
think of $A$ as a set of group labels, and $f$ as the function that assigns a
label to each element of $U$. This way, $f$ divides $U$ into non-overlapping
groups. We say that $f:U\to A$ is a \emph{partition} of $U$.

@@

\begin{algorithm}[H]
\caption{Split-Apply-Combine}
\label{alg:sac}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition $f:U\to A$}
\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs $(a,x)\in
        A\times X$ }
$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}

Here, the output is collected in a set $R$ (this set is technically a function
$A\to X$ since it pairs each $a\in A$ with one element of $X$). We also
introduced some new notation: the notation $2^U$ is the power set of $U$ and
$f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you give it a
set of labels from $A$ and it returns all the elements of $U$ that have any of
those labels. This definition is somewhat too general for the current
algorithm, but it will be necessary when we move to collapsing groups.

@@

%Collapsing schemes can be represented as tables. For example, the European NACE
%classification is a hierarchical classification of economic activity. Each
%group is encoded as a 4-digit code where the first digit indicates the highest
%grouping level, and the last digit the most detailed level. This induces a
%collapsing sequence by combining groups that have the same parent in the
%hierarchy. Below is an example, based on a small piece of the NACE code list.
%%
%\begin{equation}
%\begin{tabular}{rrr}
%\hline
% $A_0$& $A_1$ & $A_2$\\
%\hline
%0111&   011 & 01\\
%0112&   011 & 01\\
%0113&   011 & 01\\
%0121&   012 & 01\\
%0122&   012 & 01\\
%0123&   012 & 01\\
%0124&   012 & 01\\
%\hline
%\end{tabular}
%\label{eq:tablerep}
%\end{equation}
%
%Here, $A_0$ is the most detailed partition, splitting a dataset into seven
%groups, relating to $\{0111,\ldots,0124\}$. In $A_1$ there are two groups.
%The first, labeled $01.1$ is constructed by combining $0111, 0112$ and
%$0113$ and the second, labeled $01.2$ is constructed by combining
%$0121,\ldots 01.24$. Finally, in $A_2$ all groups are combined.
%
%To compute an aggregate for records labeled with $0111$, the algorithm will
%select those records and test whether the subset is suitable. If not, it will
%move to $f_1(0111)=011$ and find all the groups in $A_0$ that map to 011 as
%follows.
%\begin{displaymath}
%f_1^{-1}(\{011\}) = \{0111, 0112, 0113\}.
%\end{displaymath}
%It will then move on and find all records labeled with any of those three
%group labels, and test again. If this new data set is suitable, the result is
%computed, if not, another collapse takes place. The whole procedure is repeated 
%for each element in $A_0$.

Denote with $F_k:A_0\to A_k$, the function that  accepts a label in $A_0$ and
returns the corresponding label in $A_k$. This can formally be defined as
consecutive application of $f_1, f_2,\cdots, f_k$ to an element of $A_0$. In
notation
\begin{equation*}
F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
\end{equation*}
Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
\begin{equation*}
F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
\end{equation*}
This function asks for a set of labels in $A_k$ and returns all the labels in
$A_0$ that are mapped to those labels via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally write
down our main Algorithm~\ref{alg:saccg}.
